Linting Linearity in Core/System FC

### 2023-01-02 report.tex
Items on background sections

@@

@@

@@

@@

@@

% \begin{itemize}
%     \item Linear types
%     \item Linear Haskell
%     \item Core
%     \item Sistemas de inferência
%     \item GADTs e Coercions
% \end{itemize}

@@

>>>>

\newcommand{\mypara}[1]{\paragraph{\textbf{#1}.}}
\newcommand{\lolli}{\multimap}
\newcommand{\tensor}{\otimes}
\newcommand{\one}{\mathbf{1}}
\newcommand{\bang}{{!}}

\newcommand{\llet}[2]{\mathsf{let}\ #1\ \mathsf{in}\ #2}

@@

\mypara{Type Systems} A type system can be formally described through
a set of inference rules that inductively define a judgment of the
form $\Gamma \vdash M : A$, stating that program expression $M$ has
type $A$ according to the \emph{typing assumptions} for variables
tracked in $\Gamma$. For instance,
$x{:}\mathsf{Int}, y{:}\mathsf{Int} \vdash x+y : \mathsf{Int}$ states
that $x+y$ has type $\mathsf{Int}$ under the assumption that $x$ and
$y$ have type $\mathsf{Int}$.  An expression $M$ is deemed well-typed
with a given type $A$ if one can construct a typing derivation with $M :
A$ as its conclusion, by repeated application of the inference rules.

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{10.5555/509043} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, the term $\lambda x{:}A.x$, denoting the identity function, is a functional abstraction
taking an argument of type $A$ and returning it back.

\mypara{Linear Logic}

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.

To present the formalization of this logic, besides the new
connectives, we need to introduce the \emph{resource-aware context}
$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
also a list of variables and their types, but where each and every
variable must be used exactly once during inference.  So, to introduce
the connective $\tensor$ which defines a multiplicative pair of
propositions, we must use exactly all the resources
($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$)
proposition $A$, and ($\Delta_2$) proposition $B$:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since it
augments the form of the judgment and it's the one that ensures
logical strength i.e. we're able to translate intuitionistic logic
into linear logic.  The proposition $\bang A$ (read \emph{of course}
$A$) is used (under certain conditions) to make a resource
``infinite'' i.e. to make it useable an arbitrary number of times. To
distinguish the ``infinite'' variables, a separate, unrestricted,
context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite''
resources, and $\Delta$ the resources that can only be used once.  The
linear typing judgment for the introduction of the exponential $\bang A$
takes the form:
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \emptyset \vdash M : A}
    {\Gamma ; \emptyset \vdash \bang M : \bang A}
\]
Logically, a proof of $\bang A$ cannot use linear resources since
$\bang A$ denotes an unbounded (potentially $0$) number of copies of
$A$. Proofs of $\bang A$ may use other unrestricted or exponential
resources, tracked by context $\Gamma$.
From a computational perspective, the type $\bang A$
internalizes the simply-typed $\lambda$-calculus in the linear
$\lambda$-calculus.

The elimination form for the exponential, written $\llet{ !u
= M}{ N}$, warrants the use of resource $A$ an unbounded
number of times in $N$ via the variable $u$:
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
\]

% TODO: Revise the story above instead of dumping so much information

\mypara{Linear Types}

Girard's \emph{linear logic} suggests a new \emph{type system} for functional
languages through the Curry-Howard isomorphism -- every linear logic proposition
maps to a different linear type.

Values belonging to a linear type must be used exactly once. For example, the
type corresponding to the linear implication ($\lolli$), the linear function, is
a function that consumes its argument exactly once. A linear function that
receives an integer as an input and consumes it exactly once to
produce a character would have type $\textbf{int} \lolli
\textbf{char}$.

\begin{itemize}
    \item Ideia dos linear types
    \item Calculo lambda linear (como forma de falar da linearidade)
    \item Gestão de recursos
    \item Rust (core is linear lambda calculus?)
\end{itemize}

@@

\begin{itemize}
    \item Features fundamentais
    \item Exemplos, introduzir syntax
    \item (Ao contrário de OCaml) tem um sistema de tipos muito sufesticado
    \item Eventualmente explicação de GADTs? c/ exemplo do Vec
    \item Uma das features mais avaçadas/experimentais é o Linear Haskell
\end{itemize}

@@

\begin{itemize}
    \item Linear Haskell definition
    \item Consuming values precisely
    \item Multiplicities
    \item Multiplicity polymorphism
    \item Que relaciona o systemFC com o calculo lambda linear
\end{itemize}

@@

\begin{itemize}
    \item What is Core (IR) e as particularidades
    \item Em cima do SystemFC
    \item Optimizações feitas no Core
    \item Base SystemFC tem de ser "merged" com o linear lambda calculus
    \item Ou porque é que é importante o Core ter linearidade também/a que nível
        foi introduzido
    \item Falar das coercions em especifico
    \item E das local equalities que são implicitas no front end
\end{itemize}

@@

\begin{itemize}
    \item Parser -> Rename -> Typechecker -> Desugar
    \item Core2Core transformations
    \item GHC unique em haver tantas transformações sempre sobre a mesma intermediate
        representation como input e output
    \item 
\end{itemize}

% Previous itemize. Talvez ainda tenha de ter secções sobre os últimos dois
% pontos
% \begin{itemize}
%     \item Linear types
%     \item Linear Haskell
%     \item Core
%     \item Sistemas de inferência
%     \item GADTs e Coercions
% \end{itemize}

@@

\section{OutsideIn(X)}

Se for modificar o typechecker com as multiplicity coercions vou ter de falar
disto.
